~\Medium Increased sexual dimorphism does not evolve in a fossil stickleback following ecological release from fish piscivores

### 2024-10-21 manuscript/paper.tex
manuscript improvements, text, revisions, edits.

\title{~\Large Increased sexual dimorphism evolves in a fossil
stickleback following ecological release from fish piscivores}
\author{\large Matthew Stuart\(^{1,2}\), Allison Ozark\(^{3}\), Raheyma
Siddiqui\(^3\), Akhil Ghosh\(^{1,2}\)\\
\large Samantha Swank\(^3\), Michael A. Bell\(^4\), Gregory J.
Matthews\(^{1,2}\), and Yoel E. Stuart\(^{3,+}\)\\
\vspace{-1.1mm}\\
\large \(^1\) Department of Mathematics and Statistics, Loyola
University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^2\) Center for Data Science and Consulting, Loyola University
Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^3\) Department of Biology, Loyola University Chicago, Chicago,
IL, USA \vspace{-1.1mm}\\
\large \(^4\) University of California Museum of Paleontology, Berkeley,
CA, USA \vspace{-1.1mm}\\
\large \(^+\) Corresponding:
\href{mailto:ystuart@luc.edu}{\nolinkurl{ystuart@luc.edu}}
\vspace{-1.1mm}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Everyone loves the stickle \vspace{2mm}\\
\emph{Keywords}: Stickle
\end{abstract}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1607}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0893}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7500}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endhead
Standard Length & \texttt{stl} & Distance from anterior tip of
premaxilla to posterior end of last vertebra (hypural plate) \\
Dorsal Spine & \texttt{mds} & Number of dorsal spines from 0 to 3 \\
Dorsal Fin Ray & \texttt{mdf} & Number of bones in the dorsal fin
posterior to the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray & \texttt{maf} & Number of bones in the anal fin posterior
to the anal spine (i.e., soft dorsal fin rays) \\
Abdominal Vertebra & mav & Number of vertebrae anterior to the first
vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
Caudal Vertebra & \texttt{mcv} & Number of vertebrae posterior to and
including the first vertebra contacting an anal fin pterygiophore
(Aguirre et al.~2014) \\
Pterygiophore number & \texttt{mpt} & Number of pterygiophores anterior
to but excluding the pterygiophore under the third dorsal spine, which
is immediately anterior to and contiguous with the dorsal fin \\
Pelvic Spine length & \texttt{lps.sc} & Length from the base of one
pelvic spine above its articulation with the pelvic girdle to its distal
tip \\
Ectocoracoid & \texttt{ect.sc} & Length between the anterior and
posterior tips of the shoulder girdle base (i.e., ectocoracoid) \\
Pelvic Girdle & \texttt{tpg.sc} & Length between the anterior to
posterior tips along midline. If vestigial, the sum of longest
anterior-posterior axis for the vestiges \\
Cleithrum length & \texttt{cle.sc} & Length from free dorsal tip to
ventral tip of the cleithrum on the anterior margin of the shoulder
girdle (i.e., cleithrum) \\
Premaxilla & \texttt{pmx.sc} & Length from the anterior tip of the
premaxilla to the distal tip of the ascending process of the
premaxilla \\
Dorsal Spine & \texttt{Ds\#.sc\ \#=1/2/3} & Length from the base of a
dorsal spine above the pterygiophore to its distal tip along the
anterior edge \\
Pterygiophore & \texttt{lpt.sc} & Distance between the anterior to
posterior tips of the pterygiophore immediately preceding the 3rd dorsal
spine (when present) \\
\bottomrule()
\end{longtable}

\textcolor{red}{This table caption should be part of the table. Top row.. -YS}
Table: Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise noted.

@@

To span the gamut of stickleback diversity for our predictive imputation
model, we sampled modern stickleback from lakes containing generalist
stickleback populations (Hendry et al.
(\protect\hyperlink{ref-Hendryetal2009}{2009}); Bolnick
(\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes containing
benthic-limnetic species pairs (Baumgartner, Bell, and Weinberg
(\protect\hyperlink{ref-Baumgartneretal1988}{1988}); Schluter and
McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992})). The
generalist specimens used here were collected by YES in 2013 and were
previously described in Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}) (Table S1). These samples
were fixed in formalin, then stained for bone with Alizarin Red in 2013.
Benthic and limnetic specimens were kindly loaned by D. Schluter and S.
Blain at the University of British Columbia. The Schluter lab collected
benthic and limnetic individuals from Enos Lake in 1988 and from Emily
Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table
S1). The Enos specimens had been fixed whole in formalin and stored in
40\% isopropanol. The specimens from the other lakes were initially
preserved whole in 95\% ethanol in the field before being gradually
transferred to water then formalin in the lab and ultimately stored in
40\% isopropanol. In 2019, we stained these specimens for bone using
Alizarin Red. We next replicated fossil data collection (Table 1) on
these extant specimens. Standard length as well as pelvic-spine length
on each side were measured with calipers. We used a dissection
microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and
anal-fin rays. Right and left-side pelvic girdle lengths and
ectocoracoid lengths were measured from ventral photographs taken using
a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a
leveled Kaiser RS1 copy stand. Specimens were held in place for ventral
photographs using a small tabletop vise with an attached scale bar.
Lateral X-rays were used to measure dorsal spine length, number of
pterygiophores anterior to the pterygiophore holding the third spine,
length of the pterygiophore just anterior to the third spine, cleithrum
length, and pre-maxilla ascending branch length. We also counted
vertebrae from the X-rays: abdominal vertebrae were counted anterior to
the first vertebra with a haemal spine contacting an anal fin
pterygiophore. Caudal vertebrae were posterior, including the first
vertebra with the haemal spine contacting the anal fin pterygiophore
(following Aguirre, Walker, and Gideon
(\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the
Field Museum of Natural History. Specimens were exposed at 35kV and 4mA.
Small fish were exposed for 7s, medium fish for 8s, and large fish for
10s. We developed the film and scanned individual images of each fish
using the B\&W Negatives setting on an Epson Perfection 4990 Photo
flatbed at 2400 dpi. Measurements from photographs and X-rays were taken
with FIJI (Schindelin et al.
(\protect\hyperlink{ref-Schindelinetal2012}{2012})) and its plugin
ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}).

We dissected individuals from the generalist populations (Table S1) to
determine sex from the gonads. Individuals from the species-pair lakes
(TableS1) were previously sexed by the Schluter lab, using either
dissection or a genotyping protocol (Whom, personal communication).

The extant data used here consists of a total of 367 specimens all with
known sex (Table S1). Of these, there are 202 and 165 female and male
specimens, respectively.

@@

Briefly, We used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils. Let \(\boldsymbol{W}\) be an
\((n_{extant} + n_{fossil}) \times 1\) vector of the covariate sex of
the stickleback fish and \(\boldsymbol{Y}\) be an
\((n_{extant} + n_{fossil}) \times K\) matrix of the \(K\) phenotypes of
interest. Because the sex of the fossilized stickleback fish is
unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant sex and missing fossil sex, respectively. We imputeD
missing sex for the fossil data by sampling from the posterior
predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching, implemented in R (CITATION).
Traditionally, the choice for the number of completed dats sets is a
relatively small number such as \(M = 5\) or \(M = 10\). However, Zhou
and Reiter (\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a
larger number of imputed data sets if users intend on performing
Bayesian analysis after imputation. Therefore, we imputed \(M = 100\)
complete datasets.

\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
\textcolor{red}{THE POOLING DESCRIBED IN THE NEXT SENTENCE. WAS IT DONE, AND IF SO, HOW WAS IT USED IN DOWNSTREAM ANALYSES. MATT SAYS THAT HE RAN IS MODELS ON EACH IMPUTED DATA SET. -YS}
We then pooled all of the draws from the posterior distribution across
all the imputed data sets to estimate the posterior distributions of our
parameters of interest (i.e., the trait values), rather than using
Rubin's combining rules, following Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We used these posterior
distributions in our Bayesian analysis.

\hypertarget{modeling-trait-means-through-time-for-males-and-females}{%
\subsection{Modeling trait means through time for males and
females}\label{modeling-trait-means-through-time-for-males-and-females}}

The goal of our analysis was to estimate fossil trait means through
time, by sex, while incorporating uncertainty in the imputed data,
especially uncertainty sex classification. We took a Bayesian approach
because imputation is the process of sampling from a posterior
predictive distribution of missing data, given observed data.
\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH. -YS}

\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
To estimate fossil sample means through time, we first dropped the
extant individuals from our imputed data sets. Then, for a given imputed
dataset, we let \(W_{ij}\) be the imputed sex and
\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
\(j = 1,\ldots,n_{t}\). In addition, we denote \(Y_{K,ij}\), the last
variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of each
fish individual, a measure of body size. \begin{align}
{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

It is reasonable to assume that the other continuous traits will have
some correlation with standard length \textbf{(CITATION)}. We account
for this by adding an additional parameter, \(\gamma_k\), If
\(Y_{k,ij}\) is a continuous trait, then \begin{align}
{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

If \(Y_{k,ij}\) is a discrete trait, the conventional modelling method
is to fit a Poisson distribution. However, because the empirical fossil
data violate the assumption that the Poisson variance is equal to its
mean, we fit discrete traits to a generalized Poisson model as defined
in
(\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}}).
Specifically, if \(X \sim GP(\lambda,\alpha)\), then \begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align} where GP denotes a generalized Poisson distribution with
\(E(X) = \lambda\) and \(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\).

In addition, we allowed that two discrete traits, abdominal vertebrae
number (mav) and caudal vertebrae number (mcv) could also with standard
length (\textbf{CITATION}). This is because these are vertebral counts
that run in parallel with the standard length measurement, along the
anterior-posterior access of the fish; these vertebrae thus contribute
to the length of the fish directly, and may evolve in a correlated
fashion. If \(Y_{k,ij}\) is one of the above traits, then \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we set \(\gamma_k = 0\) because we expect
that trait counts are set during early ontogeny, and should not change
with size in adult fish. Thus, \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align} We point out that, for the discrete traits, the means are
represented by \(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\) for
ease of modeling.

>>>>

\title{~\Medium Increased sexual dimorphism does not evolve in a fossil
stickleback following ecological release from fish piscivores}
\author{\large Matthew Stuart\(^{1,2}\), Allison Ozark\(^{3}\), Raheyma
Siddiqui\(^3\), Akhil Ghosh\(^{1,2}\)\\
\large Samantha Swank\(^3\), Michael A. Bell\(^4\), Gregory J.
Matthews\(^{1,2}\), and Yoel E. Stuart\(^{3,+}\)\\
\vspace{-1.1mm}\\
\large \(^1\) Department of Mathematics and Statistics, Loyola
University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^2\) Center for Data Science and Consulting, Loyola University
Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^3\) Department of Biology, Loyola University Chicago, Chicago,
IL, USA \vspace{-1.1mm}\\
\large \(^4\) University of California Museum of Paleontology, Berkeley,
CA, USA \vspace{-1.1mm}\\
\large \(^+\) Corresponding:
\href{mailto:ystuart@luc.edu}{\nolinkurl{ystuart@luc.edu}}
\vspace{-1.1mm}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Theory suggests that populations should expand their habitat- and
resource-use niches when they are freed from interactions with
competitors and predators--so-called ecological release. One way that
this ecological release can manifest is through increased differences
between males and females of the same species as they diverge into
different niches. If divergent niche use drives corresponding divergent
adaptation in in the traits used to exploit the divergent niches, then
theory predicts an increase in sexual dimorphism for those traits,
sometimes called character displacement between the sexes. Using a
dataset of 16 traits collected from 18 temporal samples spread across
16,500 years of evolution of the fossil stickleback fish,
\emph{Gasterosteus doryssus}, we tested the prediction that release from
predation inferred in this system would result in character displacement
between the sexes. We used data from populations of extant stickleback
(\emph{Gasterosteus aculeatus}) to build an imputation model to classify
individual sex in fossil specimens. Then, we quantified the extent of
sexual dimorphism at each temporal sample, and tracked how it changed
through time. \vspace{2mm}\\
\emph{Keywords}: character displacement, character release, evolutionary
time series, \emph{Gasterosteus doryssus}, missing data imputation
\end{abstract}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2034}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0847}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7119}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endhead
Standard Length & \texttt{stl} & Distance from anterior tip of
premaxilla to posterior end of last vertebra (i.e., the llhypural
plate) \\
Dorsal Spine Count & \texttt{mds} & Number of dorsal spines \\
Dorsal Fin Ray Count & \texttt{mdf} & Number of bones in the dorsal fin
posterior to the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray Count & \texttt{maf} & Number of bones in the anal fin
posterior to the anal spine (i.e., soft anal fin rays) \\
Abdominal Vertebra Count & mav & Number of vertebrae anterior to the
first vertebra contacting an anal fin pterygiophore (Aguirre et
al.~2014) \\
Caudal Vertebra Count & \texttt{mcv} & Number of vertebrae posterior to
and including the first vertebra contacting an anal fin pterygiophore
(Aguirre et al.~2014) \\
Pterygiophore Count & \texttt{mpt} & Number of pterygiophores anterior
to but excluding the pterygiophore under the third dorsal spine \\
Pelvic Spine Length & \texttt{lps.sc} & Length from the base of one
pelvic spine at its pelvic girdle articulation to its distal tip \\
Ectocoracoid Length & \texttt{ect.sc} & Length from the anterior to
posterior tips of the shoulder girdle base \\
Pelvic Girdle Length & \texttt{tpg.sc} & Length from the anterior to
posterior tips along midline. If vestigial, the sum of longest axis
along vestiges \\
Cleithrum Length & \texttt{cle.sc} & Length from dorsal tip to ventral
tip on the anterior margin of the shoulder girdle \\
Premaxilla Length & \texttt{pmx.sc} & Length from the anterior tip of
the premaxilla to the distal tip of its ascending process \\
Dorsal Spine Length & \texttt{Ds\#.sc\ \#=1/2/3} & Length from the
dorsal spine base at the pterygiophore to its distal tip \\
Pterygiophore Length & \texttt{lpt.sc} & Distance from anterior to
posterior edges of the pterygiophore immediately before ds3 (when
present) \\
\bottomrule()
\end{longtable}

\textcolor{red}{This table caption should be part of the table. Top row. at the very least, it should be the width of the table. -YS}
Table: Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise noted.

@@

To span the gamut of stickleback diversity for our predictive imputation
model, we sampled modern stickleback from lakes containing generalist
stickleback populations (Hendry et al.
(\protect\hyperlink{ref-Hendryetal2009}{2009}); Bolnick
(\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes containing
benthic-limnetic species pairs (Baumgartner, Bell, and Weinberg
(\protect\hyperlink{ref-Baumgartneretal1988}{1988}); Schluter and
McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992})). The
generalist specimens used here were collected by YES in 2013 and were
previously described in Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}) (Table S1). These samples
were fixed in formalin, then stained for bone with Alizarin Red in 2013.
Benthic and limnetic specimens were kindly loaned by D. Schluter and S.
Blain at the University of British Columbia. The Schluter lab collected
benthic and limnetic individuals from Enos Lake in 1988 and from Emily
Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table
S1). The Enos specimens had been fixed whole in formalin and stored in
40\% isopropanol. The specimens from the other lakes were initially
preserved whole in 95\% ethanol in the field before being gradually
transferred to water then formalin in the lab and ultimately stored in
40\% isopropanol. In 2019, we stained these specimens for bone using
Alizarin Red.

We next replicated fossil data collection (Table 1) on these extant
specimens. Standard length as well as pelvic-spine length on each side
were measured with calipers. We used a dissection microscope to count
dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right
and left-side pelvic girdle lengths and ectocoracoid lengths were
measured from ventral photographs taken using a Canon EOS Rebel T7 with
a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy
stand. Specimens were held in place for ventral photographs using a
small tabletop vise with an attached scale bar. Lateral X-rays were used
to measure dorsal spine length, number of pterygiophores anterior to the
pterygiophore holding the third spine, length of the pterygiophore just
anterior to the third spine, cleithrum length, and pre-maxilla ascending
branch length. We also counted vertebrae from the X-rays: abdominal
vertebrae were counted anterior to the first vertebra with a haemal
spine contacting an anal fin pterygiophore. Caudal vertebrae were
posterior, including the first vertebra with the haemal spine contacting
the anal fin pterygiophore (following Aguirre, Walker, and Gideon
(\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the
Field Museum of Natural History. Specimens were exposed at 35kV and 4mA.
Small fish were exposed for 7s, medium fish for 8s, and large fish for
10s. We developed the film and scanned individual images of each fish
using the B\&W Negatives setting on an Epson Perfection 4990 Photo
flatbed at 2400 dpi. Measurements from photographs and X-rays were taken
with FIJI (Schindelin et al.
(\protect\hyperlink{ref-Schindelinetal2012}{2012})) and its plugin
ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}). We dissected
individuals from the generalist populations (Table S1) to determine sex
from the gonads. Individuals from the species-pair lakes (TableS1) were
previously sexed by the Schluter lab, using either dissection or a
genotyping protocol \textcolor{red}{(Whom, personal communication)}.

@@

Briefly, we used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils. Let \(\boldsymbol{W}\) be an
\((n_{extant} + n_{fossil}) \times 1\) vector of the covariate sex of
the stickleback fish and \(\boldsymbol{Y}\) be an
\((n_{extant} + n_{fossil}) \times K\) matrix of the \(K\) phenotypes of
interest. Because the sex of the fossilized stickleback fish is
unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant sex and missing fossil sex, respectively. We imputed
missing sex for the fossil data by sampling from the posterior
predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching, implemented in R (CITATION).
Traditionally, the choice for the number of completed dats sets is a
relatively small number such as \(M = 5\) or \(M = 10\). However, Zhou
and Reiter (\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a
larger number of imputed data sets if users intend on performing
Bayesian analysis after imputation. Therefore, we imputed \(M = 100\)
complete datasets.
\textcolor{red}{PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}

\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}

We then pooled all of the draws from the posterior distribution across
all the imputed data sets to estimate the posterior distributions of our
parameters of interest (i.e., the trait values), rather than using
Rubin's combining rules, following Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We then dropped extant
fish from our imputed data sets and used these posterior distributions
in our downstream analyses of fossil sexual dimorphism.

\hypertarget{modeling-fossil-stickleback-trait-means-through-time-for-males-and-females}{%
\subsection{Modeling fossil stickleback trait means through time for
males and
females}\label{modeling-fossil-stickleback-trait-means-through-time-for-males-and-females}}

The goal of our analysis was to estimate fossil trait means through
time, by sex, while incorporating uncertainty across imputations in the
classification of sex. We used a Bayesian approach to estimate trait
means in part because imputation is the process of sampling from a
posterior predictive distribution of missing data, given observed data.
\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE. -YS}

\textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS}

To estimate fossil sample means through time, for a given imputed
dataset, we let \(W_{ij}\) be the imputed sex and
\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
\(j = 1,\ldots,n_{t}\). In addition, we denote \(Y_{K,ij}\), the last
variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of each
fish individual, a measure of body size. \begin{align}
{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

Continuous traits will likely have some correlation with standard length
(i.e., allometry) (Huxley
(\protect\hyperlink{ref-Huxley1932}{1932}))(Voje, Bell, and Stuart
(\protect\hyperlink{ref-Vojeetal2022}{2022}))(Voje et al.
(\protect\hyperlink{ref-Vojeetal2014}{2014})). We account for this by
adding an additional parameter, \(\gamma_k\), If \(Y_{k,ij}\) is a
continuous trait, then \begin{align}
{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

If \(Y_{k,ij}\) is a discrete trait, the conventional modelling method
is to fit a Poisson distribution. However, the empirical fossil data
violate the assumption that the Poisson variance is equal to its mean,
so we fitted discrete traits to a generalized Poisson model
((\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}})).
Specifically, if \(X \sim GP(\lambda,\alpha)\), then \begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align} where GP denotes a generalized Poisson distribution with
\(E(X) = \lambda\) and \(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\).

In addition, we modeled two discrete traits, abdominal vertebrae number
(mav) and caudal vertebrae number (mcv) to vary with standard length.
This is because these are vertebral counts that made along the
anterior-posterior access of the fish; these vertebrae contribute to the
length of the fish directly and may evolve in a correlated fashion. If
\(Y_{k,ij}\) is one of the above traits, then \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete counting traits, we set \(\gamma_k = 0\) because
we expect that trait counts are set during early ontogeny and should not
change with size in adult fish. Thus, \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align} We point out that counting trait means are represented by
\(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\) for ease of
modeling.

In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
model the time-\(t_i\), sample-specific trait mean \(k\) for females and
males, respectively. To account for the possibility that there is a
time-dependent (i.e., evolutionary trend) trend in \(\mu_{k,ft_i}\) and
\(\mu_{k,mt_i}\), we further set \begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align} for \(g \in \{f,m\}\), where \(\beta_{0,kg}\) and
\(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
sex, and \(u_{k,gt_i}\) is the corresponding residual. To account for
potential correlations between the residuals for a given trait \(k\) and
sex \(g\), we fit an Ornstein-Uhlenback (OU) process (Uhlenbeck and
Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More specifically,
we defined \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the change in
\(u_k,gt\) for a given trait \(k\) and sex \(g\) over a miniscule time
period \(dt\). The OU process is defined as \begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align} where \(\kappa_k\) is a parameter associated with the
correlation between \(u_{k,gt}\) and \(u_{k,g(t+dt)}\), \(\tau_k\) is
the standard deviation of the OU process, and \(W_t\) is a standard
Brownian motion. As shown in Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930}), the closed form solution for
the stochastic differential equation in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \overset{iid}{\sim}\mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align} for \(i = 2,\cdots,T\). In a traditional OU process, the
initial value \(u_{k,gt_1}\) is assumed to be a (potentially unknown)
constant. This unknown constant can therefore be considered a latent
variable, further justifying our Bayesian approach, which better
incorporates prior information and uncertainty from missing data
imputation in the estimate of latent variables.

\textcolor{red}{Q FOR MATT. Is transforming the GP onto the real number line separate from setting the priors? If so, should the statement of transformation go in the section where the GP and discrete trait modeling is discussed. Then priors get their own section here? -YS}
To aid in the sampling procedure for the discrete phenotypes, we first
had to transform the parameters of the generalized Poisson distribution
onto the real number line in order to ensure our algorithm can properly
sample from the posterior distribution of interest using STAN software.
This was done by declaring
\(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
where
\(\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.\).
